---
output:
  pdf_document: default
  html_document: default
---
# Section 1: What we measure and why 

## Mammaprint Gene Signature
* Exploring genes used in the Mammaprint gene signature - assess risk of breast cancer 
* Diagnostic signature using gene expression levels of 70 genes
* Information about the 70 gene signature used in the Mammaprint algorithm 

```{r}
library(genefu)
data(sig.gene70)
dim(sig.gene70)
head(sig.gene70)[,1:6]
```

## Assessment: Phenotypes 
* COPDSexualDimorphism.data package - phenotypes (cols) individuals (rows)
* Data to assess incidence of COPD and emphysema by gender and smoking status
* The pkyrs variable in the expr.meta data.frame represents pack years smoked. Other variables include gender and diagmaj (disease status). These variables correspond to phenotypes.

```{r}
library(COPDSexualDimorphism.data)
data(lgrc.expr.meta)
```

## Assessment: Chromosomes and SNPs 
* GWAS (Genome-wide association studies)
* Comparing individuals with disease vs. controls using SNP chips or DNA sequencing. 
* SNPs with association are investigated for disruption of gene regulation or function 
* Bioconductor *gwascat* package

```{r}
library(gwascat)
data(ebicat_2020_04_30)
ebicat_2020_04_30
```

## Microarray Technology 1: How Hybridization Works 
* Two technologies: microarray and NGS 
* Both counting DNA or RNA molecules 
* Both use a trick which allows us to take double-stranded DNA and convert to single-stranded 
* Both require thousands - millions of molecules for us to be able to measure anything 
* If a few cells only, they must be amplified 

### Microarray Technology 
1. Denaturation (single-stranded)
2. Hybridization - when you have a single strand in solution and it finds complimentary DNA, it will hybridize to form 2 stranded DNA. This can be exploited to count molecules 
3. Can create probes / troughs for different sequences. Put on location on piece of solid for the molecules we want to be able to count. Probes have compliments to the DNA that we want to count. 

### Labeling 
Need indirect ways to count molecules. Labeling adds a chemical to each molecule, use optical scanner to identify the different intensities based on # labels and quantify. 

Design attribute of different technologies: synthetically sequenced, or cloned. Densities of probes put on the solid is also variable across different technologies. Also # samples on each array differs. 
Major manufacturers: 
1. Affymetrix (high density, one color)
2. Agilent (circles on grid, one or two color)
3. Illumina (high density, one or two color)
   * Uses beads instead of in-situ sequencing
