---
title: "PH525.5x Section 4: Genomic annotation with Bioconductor"
author: "Lauren Kemperman"
date: "4/13/2021"
output: pdf_document
---

## Representing Reference Sequence 
* Annotation concept hierarchy 
* Base - reference genomic sequence for an organism 
* Above this, organize the chromosomal sequence into regions of interest - i.e. 
genes, transcripts 
* SNPs and CpG sites are also regions of interest 
* SNPS are single nucleotide 
* Other varients -- indels, structural variants, fusions can constitute regions of interest but are more complicated to express + represent 
* Within ROI, identify platform oriented annotation provided by assay manufacturer 
* Once manufacturing happens, genomic annotation proceeds and annotations must be updated to account for ambiguities or updates for assay probe elements 
* Above genomic sequence ROIs, annotations concerning groups with shared structural or functional properties 
* Pathways with nodes being genes and paths being relationships between gene products, i.e. protein protein interaction, promotion, enhancement, repression (3rd level of hierarchy)
* Begin with reference genomes 
* Biostrings package - **available.genomes** - packages that represent reference genomic sequences for many different organisms 
* Homo sapiens reference - some have repeat masking and there are versions which include the masked regions 
  
  - different numbers of sequences in the two builds due to contigs that haven't been placed on chromosomes yet 

* Operations defined for BSGenome objects - substring, extract chromosomal information 
* Bases in full sequence aren't completely resolved 
* Application of iteration - count the number of bases in a number of chromosomes

* If you have enough RAM, it is possible to operate on chromosomes in parallel and performing operations using multicore programming 
```{r}
library(BSgenome)
library(Biostrings)
ag = available.genomes()
grep("Scerev", ag, value=TRUE)
grep("Hsap", ag, value=TRUE)

# inspect the human genome
library(BSgenome.Hsapiens.UCSC.hg19)
Hsapiens
length(Hsapiens)
class(Hsapiens)
methods(class="BSgenome")

# inspect human genome
Hsapiens$chrX
substr(Hsapiens$chrX, 5e6, 5.1e6)
nchar(Hsapiens$chrY)
nchar(Hsapiens[[24]])

library(parallel)
options(mc.cores=detectCores())

system.time(sum(unlist(mclapply(18:24, function(x) nchar(Hsapiens[[x]])))))
```

## Assessment: Reference Genomes


```{r}

library(BSgenome)
library(Biostrings)
ag = available.genomes()
library(BSgenome)
grep("mask", grep("Drerio", available.genomes(), value=TRUE), invert=TRUE, value=TRUE) # exclude masked


library(BSgenome.Hsapiens.UCSC.hg19.masked)
c17m = BSgenome.Hsapiens.UCSC.hg19.masked$chr17

c22m = BSgenome.Hsapiens.UCSC.hg19.masked$chr22
round(100*sum(width(masks(c22m)$AGAPS))/length(c22m),0)

```

## Gene, Transcript and Exon Databases 

* Can find information about reference genome regions such as genes, transcripts and exons on annotation packages 
* UCSC Genome Browser - major source of reference genome structure annotation 
* **TxDb.Hsapiens.UCSC.hg19** - collection of well documented protein coding genes, transcripts and exons on the hg19 build of the human genome. Additional TxDb packages exist for other organisms and genome builds 
* Introduction to TxDb package architecture 

```{r}
# Import TxDb transcript database
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb = TxDb.Hsapiens.UCSC.hg19.knownGene
class(txdb)
methods(class="TxDb")

# extract and inspect genes from TxDb 
genes(txdb)
table(strand(genes(txdb)))
summary(width(genes(txdb)))

# inspect larges gene in genome 
id = which.max(width(genes(txdb)))
genes(txdb)[id]
library(org.Hs.eg.db)
select(org.Hs.eg.db, keys="286297", keytype="ENTREZID", columns=c("SYMBOL", "GENENAME"))

# compare total size of exons to total size of genes 
ex = exons(txdb)
rex = reduce(ex)
ex_width = sum(width(rex)) # bases in exons 
gene_width = sum(width(genes(txdb))) # bases in genes 
ex_width/gene_width
```

## ensembldb, EnsDb: annotation from EMBL

* European initiative for annotating genome called ensembl 
* Ensemble-based representations managed in package called EmsembleDb 
* Different packages representing different builds of ensembl annotation for different organisms 
* More direct relationship to database and database tables - gene, transcipt, transcript to exon mapping tables. 
* More details provided to user through Ensembl transcripts method - get info on transcripts but also associated proteins, genes and biotype 


```{r}
# inspect data available from Ensembl
library(ensembldb)
library(EnsDb.Hsapiens.v75)
names(listTables(EnsDb.Hsapiens.v75))

# extract Ensembl transcripts 
edb = EnsDb.Hsapiens.v75 # abbreviate 
txs <- transcripts(edb, filter = GeneNameFilter("ZBTB16"), 
                   columns = c("protein_id", "uniprot_id", "tx_biotype"))
txs 

# compare Ensembl and UCSC transcripts 
alltx = transcripts(edb) # Ensembl is larger 
utx = transcripts(txdb) # UCSC is smaller 

# table of biological types of transcripts 
table(alltx$tx_biotype)
```
## Assessment: Gene and transcript model

```{r}
library(devtools)
install_github("genomicsclass/ph525x")
library(ph525x)
stopifnot(packageVersion("ph525x") >= "0.0.16") # do over if fail 
modPlot("ESR1", useGeneSym=FALSE, collapse=FALSE) 

library(TxDb.Hsapiens.UCSC.hg19.knownGene)
txdb = TxDb.Hsapiens.UCSC.hg19.knownGene
e_id <- select(edb, keys="ESR1", keytype="GENENAME", columns=c("ENTREZID"))[1, "ENTREZID"]
n_transcripts <- length(transcripts(txdb, filter=list(gene_id=e_id)))
paste("Number of transcripts comprimising model of ESR1: ", n_transcripts)
```

## AnnotationHub: finding and caching important information 

* Central hub for genomic annotation files maintained by Bioconductor community 
* Includes annotation files from UCSC, ENSEMBL, and the Broad Institute 
* **AnnotationHub** allows you to search and download resources from inside R session 

```{r}
library(AnnotationHub)
ah <- AnnotationHub()
ah 

length(unique(ah$species))

ah_human <- subset(ah, species == "Homo sapiens")
ah_human 

query(ah, "HepG2")
query(ah, c("HepG2", "H3K4me3"))
hepg2 <- query(ah, "HepG2")
hepg2_h3k4me3 <- query(hepg2, c("H3k4me3"))
hepg2_h3k4me3
hepg2_h3k4me3$tags 

# display(query(ah, "HepG2"))

e118_broadpeak <- query(hepg2_h3k4me3, c("E118", "broadPeak"))
id <- e118_broadpeak$ah_id 
id 

hepg2_h3k4me3_broad <- ah[["AH29728"]]
hepg2_h3k4me3_broad
alt_format <- ah[[id]]
identical(hepg2_h3k4me3_broad, alt_format)

```
## Assessment: AnnotationHub

```{r}
library(AnnotationHub)
ah = AnnotationHub()
mah = mcols(ah)
names(mah)
sort(table(mah$species), decreasing=TRUE)[1:10]

n_ctcf_binding_hepg2 <- length(names(query(query(ah, "HepG2"), "CTCF")))
paste("Number of entries addressing CTCF binding in HepG2: ", n_ctcf_binding_hepg2)
```
## liftOver: Translating between reference builds 

* Genomic annotations typically defined for fixed genome build 
* Human is often hg19 
* When analysis is performed on different genome build, annotations must be translated to the coordinates of the new build before use 
* Process of translating called **lifting** 
* Implemented in **liftOver()** function of **rtracklayer** Bioconductor package 
* Tutorial will move features from genome build hg38 -> hg19 

```{r}
# liftOver from rtracklayer 
library(rtracklayer)
?liftOver 

# chromosome 1 gene locations in hg38 
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
tx38 <- TxDb.Hsapiens.UCSC.hg38.knownGene
seqlevels(tx38, pruning.mode="coarse") = "chr1"
g1_38 <- genes(tx38)

# Download hg38 to hg19 chain file 
library(AnnotationHub)
ah <- AnnotationHub()
ah.chain <- subset(ah, rdataclass == "ChainFile" & species == "Homo sapiens")
query(ah.chain, c("hg19", "hg38"))
ch <- ah [["AH14108"]]

# perform the liftOver 
g1_19L <- liftOver(g1_38, ch)
g1_19L

```
## Assessment: liftOver 

```{r}
if(!file.exists("hg19ToHg38.over.chain")){
  download.file("http://hgdownload.cse.ucsc.edu/goldenPath/hg19/liftOver/hg19ToHg38.over.chain.gz", "hg19ToHg38.over.chain.gz")
  library(R.utils)
  gunzip("hg19ToHg38.over.chain.gz")
}

library(ERBS)
data(HepG2)
library(rtracklayer)
ch = import.chain("hg19ToHg38.over.chain") 
nHepG2 = liftOver(HepG2, ch)

s1 <- start(HepG2[1])
s2 <- start(nHepG2[1])[[1]]

abs_diff_bases <- abs(s2 - s1)
paste("Number of bases moved upstream in first range of HepG2 to hg38: ", abs_diff_bases)

```

* **rtracklayer** package parses data into common formats so they can easily be used as annotations in future analysis 

```{r}
library(devtools)
install_github("genomicsclass/ERBS") # install ERBS package 

f1 = dir(system.file("extdata",package="ERBS"), full=TRUE)[1] # access dat a
readLines(f1, 4) # preview a few lines 

library(rtracklayer)
imp = import(f1, format="bedGraph") # import as bedGraph format
imp

genome(imp) # genome identifier tag not set, but can be set manually 
genome(imp) = "hg19"
genome(imp)

export(imp, "demoex.bed") # export as BED format
cat(readLines("demoex.bed", n=5), sep="\n") # check output file
```
## Assessment: Import/export 

```{r}

library(rtracklayer)
data(targets)
c_targets <- class(targets)
paste("Class of targets: ", c_targets)

library(GenomicRanges)
mtar <- with(targets,
GRanges(chrom, IRanges(start,end), strand=strand,
targets=target, mirname=name))

cat(export(mtar[1:5], format="bed"), sep="\n")
cat("\n")
cat(export(mtar[1:5], format="gff3"), sep="\n")
```
## OrgDb: unified organism-specific annotation for systems biology 

* Approach to annotation in Bioconductor 
* Org packages have the form Org - two letter abbreviation of organism 
* Two-letter abbreviation of organization - entrezgene is the resource 
* Sqlite-based package 
* If you know something about a RefSeq or UniProt ID you can learn about what genes have been annotated to it 

```{r}
# load human OrgDb and inspect available keys
library(org.Hs.eg.db)
org.Hs.eg.db
keytypes(org.Hs.eg.db)

# load GO.db and inspect available terms 
library(GO.db)
allterms = keys(GO.db, keytype="TERM")
allterms[1:5]

# find GOID (gene ontology tag) for ribosome biogenesis
goid <- select(GO.db,  keys = "ribosome biogenesis", keytype="TERM", columns="GOID")[,"GOID"]

# find symbols for genes involved in ribosome biogenesis 
select(org.Hs.eg.db, keys=goid, keytype="GO", columns="SYMBOL")

# you can pull out multiple columns at once
e_id <- select(org.Hs.eg.db, keys = "GO:0042254", keytype="GO", columns=c("SYMBOL", "ENTREZID"))

entrezid <- unlist(e_id[e_id["SYMBOL"] == "ZNF658", "ENTREZID"])

# find gene ontology tags for related to ZNF658, which has the specified ENTREZID
select(org.Hs.eg.db, keys=entrezid, keytype="ENTREZID", columns="GO")

# save GO tags to a character vector
select(org.Hs.eg.db, keys=entrezid, keytype="ENTREZID", columns="GO")$"GO"
myk = unlist(.Last.value)

# identify biological processes ZNF658 is involved in
#select(GO.db, keys=myk, columns="TERM")

```
## Assessment: orgDb 

```{r}
library(org.Hs.eg.db)
org.Hs.eg.db
keytypes(org.Hs.eg.db)

genes <- select(org.Hs.eg.db, key="17q21.1", keytype="MAP", columns=c("GENENAME", "ENTREZID"))
num_genes <- nrow(genes)
paste("Number of genes present on 17q21.1: ", num_genes)

genes_with_go_tag <- select(org.Hs.eg.db, key="17q21.1", keytype="MAP", columns=c("GENENAME", "ENTREZID", "GO"))

library(plyr)

counts <- count(genes_with_go_tag, "GO")
top_five_go_id <- counts[order(-counts$freq),][1:5,]

go_annotations_for_ormdl3 <- select(org.Hs.eg.db, key="ORMDL3", keytype="SYMBOL", columns=c("EVIDENCE", "GO"))

num_tas_evidence_codes <- length(go_annotations_for_ormdl3[go_annotations_for_ormdl3$EVIDENCE == "TAS",])


paste("number of GO annotations for ORMDL3 having TAS (traceable author statement) as their evidence code: ", num_tas_evidence_codes)

```